Demographic inference using the SFS with \moments and \demes

### 2025-04-01 chapter/paper.tex
add finished draft

\usepackage{fullpage}
\usepackage{url}
\usepackage{authblk}
\usepackage{graphicx}
\usepackage{color}

@@

\begin{abstract}
In this chapter we show how to use \moments, a Python library for computing the
joint SFS and inferring demographic history, through its \demes interface. 
Following a discussion of SFS inference, we illustrate the general usage of 
\moments and sketch two specific examples. We conclude with a discussion of some 
additional considerations regarding the \moments library.
\end{abstract}

@@

Specifying demographic models requires defining populations (or ``demes'') and
their relationships via splits and gene flow. \demes provides a standardized
and accessible format for defining demographic models \citep{gower2022demes},
and has been adopted by widely used software for population genetics simulation
and inference (including \msprime \citep{baumdicker2022efficient}, \momi
\citep{dilber2024faster}, \fwdpy \citep{thornton2019polygenic}, \texttt{GADMA}
\citep{noskova2023gadma2}). \demes encourages interoperability and reuse of
code across software, ease of model implementation and reduction of errors.

Demographic models are specified in YAML format (see \citet{gower2022demes} and
the associated documentation). The two-population split-with-migration model
depicted in Figure~\ref{fig:im}A is specified as below:
\begin{minted}[gobble=4, frame=single,]{yaml}
    description: split-with-migration model, loosely based on example 2
    generation_time: 29
    time_units: years
    demes:
    - name: ancestral
      epochs:
      - {start_size: 15000, end_time: 300000}
      - {start_size: 25000, end_time: 75000}
    - name: popA
      ancestors: [ancestral]
      epochs:
      - {start_size: 30000}
    - name: popB
      ancestors: [ancestral]
      epochs:
      - {start_size: 1200, end_size: 14000}
    migrations:
    - demes: [popA, popB]
      rate: 5e-5
\end{minted}

In \moments, a \demes-specified demographic model can be directly used to
compute either the SFS or multi-population LD statistics
\citep{ragsdale2019models, ragsdale2020unbiased}. For the SFS, we simply need
to load the demographic model using \demes and specify the number of haploid
samples to draw and from which populations.
\begin{minted}[gobble=4, frame=single]{python}
    import demes, moments
    g = demes.load('model.yaml')
    samples = {'popA': 60, 'popB': 60}
    fs = moments.Demes.SFS(g, samples=samples)
\end{minted}
Here, we have not specified a mutation rate. The default behavior sets $u=1$,
so that $\theta = 4N_eu=4N_e$, where $N_e$ is the ancestral population's
initial size. With a known mutation rate, the SFS will be properly scaled by
passing the mutation rate as a keyword argument: \texttt{moments.Demes.SFS(g,
samples=samples, u=u)}. This is equivalent to scaling the output SFS above by
multiplying by \texttt{u}, as \texttt{u*fs}.

@@

Below, we show two examples of multi-population demographic inference using the
joint SFS. The first example simulates data with known demographic parameters,
which we try to reinfer using the both the original model and simpler misspecified 
ones. In the second example, we infer a relatively simple human-Neanderthal
history using data from two human populations and a high-coverage ancient
sample from the Neanderthal lineage. For both of these examples, we briefly
describe data processing and inference setup, emphasizing the high-level
components of the analyses. For detailed information about each example,
including managing data, specifying models, performing inference, computing
confidence intervals and visualizing fits, we refer readers to the accompanying
GitHub repository
(\url{https://github.com/StatisticalPopulationGenomics-2ndEd/moments}), which
is maintained with up-to-date versioning.

\begin{figure}[t!]
    \includegraphics{../example1/fig1.pdf}
    \caption{Caption placeholder}
    \label{fig:im}
\end{figure}

In order to optimize parameters in a \demes model, we need a way to specify
which parameters should be fit. For this, we use a separate YAML-formatted file
to define \texttt{parameters}, each of which points to desired value(s) in the
demographic model and sets lower and upper bounds. This file also specifies any
relative \texttt{constraints} between pairs of parameters, to ensure the
optimization routine only explores valid model space. Below, we have included
a truncated parameters file -- the full example is available on GitHub.
\begin{minted}[gobble=4, frame=single]{YAML}
    parameters:
    - name: T0
      values:
      - demes:
          ancestral:
            epochs:
              0: end_time
      lower_bound: 0
      upper_bound: 2e5
    - name: T1
      values:
      - demes:
          ancestral:
            epochs:
              1: end_time
      lower_bound: 0
      upper_bound: 1e6
    - name: Ne
      values:
      - demes:
          ancestral:
            epochs:
              0: start_size
      lower_bound: 1e2
      upper_bound: 1e6
    ...
    constraints:
    - params: [T0, T1]
      constraint: greater_than
\end{minted}

Taking the demographic model, parameters file, and observed SFS together, we
can perform inference using the \texttt{moments.Demes.Inference.optimize(...)}
function.  The initial parameter guesses are given by the input demographic
model, which may be perturbed using a keyword argument. Any value in the
demographic model that is not specified in the parameters file will remain
unchanged, and is therefore treated as a fixed parameter. Here, we also assume
we have an estimate for the total mutation rate, $U$.
\begin{minted}[gobble=4, frame=single]{python}
    import moments

@@

Confidence intervals may also be computed using 
\texttt{moments.Demes.Inference.uncerts(...)}, once a (locally) optimal model 
has been found. The following will emit a tab-separated file containing
parameter names, best-fit values and estimated standard errors.
\begin{minted}[gobble=4, frame=single]{python}
    log_file = 'log.txt'

    uncerts = moments.Demes.Inference.uncerts(
      output, params_file, data,
      uL=U, output=log_file
    )
\end{minted}

\subsection*{Inferring parameters in a simulated split-with-migration model}

@@

When fitting the two misspecified models that do not allow for size changes in
the ancestral population, this split time is inferred to be substantially
deeper in the past than the true split time. This effect, as demonstrated by
\citet{momigliano2020biases}, is due to the decreased coalescence rate within
the ancestral population between the time of the expansion and divergence of
the descendent populations. These two misspecified models
(Figure~\ref{fig:im}H,I) also provide a worse fit to the data, as seen by the
lower log-likelihoods and the large residuals between model predictions and
data (Figure~\ref{fig:im}E-G).

\subsection*{Inferring human-Neanderthal demographic parameters}

We obtained modern human genome sequences from a recent resequencing of the 
1000 Genomes Project cohort \citep{byrska2022high}, and used the high-coverage 
sequence of the Vindija Neandertal from \cite{prufer2017high}. 
From the 1000 Genomes cohort we took 85 MSL (Mende from Sierre Leone) and 91 
GBR (British from England and Scotland) sequences as our modern human samples.  
We filtered out sites that lay outside the 1000 Genomes `strict' mask, 
were not genotyped in all four existing high-coverage archaic genome sequences, 
thatlacked high-confidence ancestral state assignments, or fell within 10 kb of 
exonic or promoter regions and were left with approximately 960 Mb of 
well-characterized and putatively neutrally-evolving sequence.
We estimated the genome-wide SFS using built-in \moments parsing functions.
Details concerning data processing (including our liftover of the Vindija 
sequence to a more recent genome build) can be found in the linked GitHub 
repository.

The model that we wished to fit is relatively simple but still has many 
parameters (11 in the example we present), so we constructed it in stages by 
first fitting simpler one-- and two--population models.
Throughout, we used $u=1.5\cdot 10^{-8}$ as an estimate of the mutation rate per 
generation and 29 years as the average generation time. We began by fitting 
one-population models with ancestral and recent size changes for MSL and GBR.
For the MSL model, we allowed three epochs with constant effective size in each 
epoch. We attributed to GBR a sharp contraction $\sim$60 kya followed by 
exponential growth to represent the out-of-Africa bottleneck and subsequent 
rapid expansion. YAML files representing the models and parameters in this 
section can be found in the linked GitHub repository.

We next stitched these one-population models together and refitted the 
resulting graph to get a two-population model that included continuous symmetric 
migration between MSL and GBR and an ancestral size expansion. Following this,
we added a Neandertal branch and fitted the 
Neandertal-modern human divergence time, Neandertal effective size, and the 
Neandertal-to-GBR admixture proportion. We also fitted the divergence time 
between the sampled Vindija Neandertal lineage and the Neandertal lineage which 
admixed with modern non-African humans. Finding that this parameter tended to 
converge to its lower bound (imposed by the estimated time of death of the 
Vindija Neandertal, ~55 kya), we fixed it to a value supported by literature 
(90 kya, \cite{prufer2017high}) in a second round of optimization.
Rather than fitting it, we fixed the time of the Neandertal-GBR pulse to 50 kya.
Fixing parameters for which good estimates are available in the literature 
(see e.g. \cite{sumer2025earliest}) may be reasonable in the first stages of 
model optimization, and we expected this value to be especially difficult to fit 
due to the small pulse proportion and Neandertal sample size. In the second 
round of optimization on the three-population model, we refit all parameters 
aside from the two event times mentioned above to refine the estimates.

After this refitting step, we were left with parameters in good agreement with 
prior authors, with a Neandertal-modern human divergence time of 547 kya, 
MSL-GBR split time of 77 kya, and Neandertal-GBR pulse proportion of 2.7 
percent. We observe some large residuals in our best-fit model 
Figure~\ref{fig:humans}G-I), suggesting ...

\begin{figure}[t!]
    \includegraphics{../example2/fig2.pdf}
    \caption{
      A plot of the best-fit model created using \demesdraw is displayed in (A).
      Marginal SFS fits between the model and empirical data are shown for MSL 
      (B) and GBR (C).
      Marginal two-way spectra predicted by the model and their residuals 
      against the data are plotted for the three pairs of populations MSL, GBR 
      (D and G), MSL, Vindija (E and H), and GBR, Vindija (F, I).
    }
    \label{fig:humans}
\end{figure}

@@

\begin{enumerate}
    \item Include here general ideas about the strengths and weaknesses of
        various approaches in population genetic inference, including when
        using the SFS.
    \item Challenges in finding local/global optima.
    \item Challenges in exploring parameter space.
    \item Background selection [Ewing, Johri]
    \item Gene dense vs gene sparse genomic architectures among species.
\end{enumerate}

>>>>

\usepackage{fullpage}
\usepackage{url}
\usepackage{authblk}
\usepackage{graphicx}
\usepackage{color}
\usepackage{setspace}

@@

%\onehalfspacing

\begin{abstract}

    The site frequency spectrum (SFS) describes the distribution of allele
    frequencies in one or more populations and sees wide use in evolutionary
    inference. Here we show how to use \moments, a Python library for working
    with the SFS and other measures of genetic diversity, and \demes, a
    standard format for parameterizing population models, to infer demographic
    history. Following an introduction to SFS inference, we illustrate the
    general usage of \moments and its \demes interface, and then show two
    specific examples, using both simulated and human genetic data. We conclude
    with caveats and considerations regarding the \moments library and SFS
    inference more broadly.

\end{abstract}

@@

Specifying demographic models requires defining populations (or ``demes'') and
their relationships via splits and gene flow. \demes provides a standardized
and accessible format for defining demographic models \citep{gower2022demes},
and has been adopted by widely used software for population genetics simulation
and inference (including \msprime \citep{baumdicker2022efficient}, \momi
\citep{dilber2024faster}, \fwdpy \citep{thornton2019polygenic}, and
\texttt{GADMA} \citep{noskova2023gadma2}). \demes encourages interoperability
and reuse of code across software, ease of model implementation and reduction
of errors.

Demographic models are specified in YAML format (see \citet{gower2022demes} and
the associated documentation). To illustrate, the two-population
split-with-migration model depicted in Figure~\ref{fig:im}A is specified as
below:
\begin{minted}[gobble=4, frame=single,]{yaml}
    description: split-with-migration model, loosely based on example 2
    generation_time: 29
    time_units: years
    demes:
    - name: ancestral
      epochs:
      - {start_size: 15000, end_time: 300000}
      - {start_size: 25000, end_time: 75000}
    - name: popA
      ancestors: [ancestral]
      epochs:
      - {start_size: 30000}
    - name: popB
      ancestors: [ancestral]
      epochs:
      - {start_size: 1200, end_size: 14000}
    migrations:
    - demes: [popA, popB]
      rate: 5e-5
\end{minted}

In \moments, a \demes-specified demographic model can be directly used to
compute either the SFS or multi-population LD statistics
\citep{ragsdale2019models, ragsdale2020unbiased}. For the SFS, we simply need
to load the demographic model using \demes and then specify the number of
haploid samples to draw and from which populations.
\begin{minted}[gobble=4, frame=single]{python}
    import demes, moments
    g = demes.load('model.yaml')
    samples = {'popA': 60, 'popB': 60}
    fs = moments.Demes.SFS(g, samples=samples)
\end{minted}
Here, we have not specified a mutation rate. The default behavior sets $u=1$,
so that $\theta = 4N_eu=4N_e$, where $N_e$ is the ancestral population's
initial size. With a known mutation rate, the SFS will be properly scaled by
passing the mutation rate as a keyword argument: \texttt{moments.Demes.SFS(g,
samples=samples, u=u)}. This is equivalent to scaling the output SFS above by
multiplying by \texttt{u}, as \texttt{u*fs}.

@@

Below, we show two examples of multi-population demographic inference using the
joint SFS. The first example simulates data with known demographic parameters,
which we try to reinfer using the both the original model and simpler
misspecified models. In the second example, we infer a relatively simple
human-Neanderthal history using data from two human populations and a
high-coverage ancient sample from the Neanderthal lineage. For both of these
examples, we briefly describe data processing and inference setup, emphasizing
the high-level components of the analyses. For detailed information about each
example, including managing data, specifying models, performing inference,
computing confidence intervals and visualizing fits, we refer readers to the
accompanying GitHub repository
(\url{https://github.com/StatisticalPopulationGenomics-2ndEd/moments}), which
is maintained with up-to-date versioning.

In order to optimize parameters in a \demes model, we need a way to specify
which parameters should be fit. For this, we use a separate YAML-formatted file
to define \texttt{parameters}, each of which points to desired value(s) in the
demographic model and for which lower and upper bounds may be set. This file
also specifies any relative \texttt{constraints} between pairs of parameters,
to ensure the optimization routine only explores valid model space. Below, we
have included a truncated parameters file -- the full example is available on
GitHub (URL above).
\begin{minted}[gobble=4, frame=single]{YAML}
    parameters:
    - name: T0
      values:
      - demes:
          ancestral:
            epochs:
              0: end_time
      lower_bound: 0
      upper_bound: 2e5
    - name: T1
      values:
      - demes:
          ancestral:
            epochs:
              1: end_time
      lower_bound: 0
      upper_bound: 1e6
    - name: Ne
      values:
      - demes:
          ancestral:
            epochs:
              0: start_size
      lower_bound: 1e2
      upper_bound: 1e6
    ...
    constraints:
    - params: [T0, T1]
      constraint: greater_than
\end{minted}

Taking the demographic model, parameters file, and observed SFS together, we
can perform inference using the \texttt{moments.Demes.Inference.optimize()}
function. The initial parameter guesses are given by the input demographic
model, which may be perturbed using a keyword argument. Any value in the
demographic model that is not specified in the parameters file will remain
unchanged, and is therefore treated as a fixed parameter. Here, we also assume
we have an estimate for the total mutation rate, $U$.
\begin{minted}[gobble=4, frame=single]{python}
    import moments

@@

Confidence intervals may also be computed using
\texttt{moments.Demes.Inference.uncerts()}, once a (locally) optimal model has
been found. The following will emit a tab-separated file containing parameter
names, best-fit values and estimated standard errors.
\begin{minted}[gobble=4, frame=single]{python}
    log_file = 'log.txt'

    uncerts = moments.Demes.Inference.uncerts(
        output, params_file, data,
        uL=U, output=log_file
    )
\end{minted}
Uncertainties may be computed using either the Fisher Information Matrix, which
tends to underestimate confidence intervals due to nonindependence of linked
sites, or using the Godambe Information Matrix, which uses a set of
bootstrapped datasets to correct for this nonindependence in the data
\citep{coffman2016computationally}. Example 1 in the associated GitHub
repository demonstrates this using a simulated dataset.

\subsection*{Example 1: Inferring parameters in a simulated split-with-migration model}

\begin{figure}[t!]
    \includegraphics{../example1/fig1.pdf}
    \caption{
        Inference using simulated data. (A) The true demographic model,
        including an expansion in the ancestral population and symmetric
        migration between the descendent populations. (B--D) The simulated
        data, showing both marginal and the joint SFS. (E--G) Residuals
        between the data SFS (in D) and the predicted SFS from the inferred
        demographic models in (H--J), resp.  The two misspecified models (H and
        I) provide worse fits to the data and both overestimate the split time
        due to disallowing any size change in the ancestral population.
    }
    \label{fig:im}
\end{figure}

@@

When fitting the two misspecified models that do not allow for size changes in
the ancestral population, this split time is inferred to be substantially
deeper in the past than the true split time. This effect, as demonstrated by
\citet{momigliano2021biases}, is due to the decreased coalescence rate within
the ancestral population between the time of the expansion and divergence of
the descendent populations. These two misspecified models
(Figure~\ref{fig:im}H,I) also provide a worse fit to the data, as seen by the
lower log-likelihoods and the large residuals between model predictions and
data (Figure~\ref{fig:im}E-G).

\subsection*{Example 2: Inferring human-Neanderthal demographic parameters}

This example uses publicly available data to infer a demographic model for two
human groups (a northern European and western African population) and a
Neanderthal individual. As shown below, even with just a single sample from the
Neanderthal lineage, we are able to recover key parameters such as the
human-Neanderthal divergence and introgression proportion.

We obtained modern human genome sequences from a recent resequencing of the 
1000 Genomes Project cohort \citep{byrska2022high}, and used the high-coverage 
sequence of the Vindija Neandertal from \cite{prufer2017high}. 
From the 1000 Genomes cohort we took 85 MSL (Mende from Sierre Leone) and 91 
GBR (British from England and Scotland) sequences as our modern human samples.  
We filtered out sites that lay outside the 1000 Genomes `strict' mask, 
were not genotyped in the high-coverage archaic genome sequences, 
that lacked high-confidence ancestral state assignments, or fell within 10 kb of 
exonic or promoter regions. This left us with approximately 960 Mb of 
well-characterized and putatively neutrally-evolving sequence.
We estimated the genome-wide SFS using built-in \moments parsing functions.
Details concerning data processing (including our liftover of the Vindija 
sequence to a more recent genome build) can be found in the linked GitHub 
repository.

\begin{figure}[t!]
    \includegraphics{../example2/fig2.pdf}
    \caption{
        Inference using human and Neanderthal data.
        (A) The best-fit model including the Vindija Neanderthal and two human
        populations.
        (B, C) Marginal SFS fits between the model and empirical data are shown
        for MSL and GBR.
        (D--I) Marginal two-way spectra predicted by the model and their residuals 
        against the data are plotted for the three pairs of populations MSL, GBR 
        (D and G), MSL, Vindija (E and H), and GBR, Vindija (F, I).
    }
    \label{fig:humans}
\end{figure}

The model that we wished to fit is relatively simple but still has many 
parameters (11 in the example we present), so we constructed it in stages by 
first fitting simpler one-- and two--population models.
Throughout, we used $u=1.5\times 10^{-8}$ as an estimate of the mutation rate per 
generation and 29 years as the average generation time. When the goal is to infer
a complex demographic history involving multiple populations, we typically start
simple, with subsets of the full data, and sequentially add complexity.
Here, we began by fitting 
one-population models with ancestral and recent size changes for MSL and GBR.
For the MSL model, we allowed three epochs with constant effective size in each 
epoch. We attributed to GBR a sharp contraction $\sim$60 kya followed by 
exponential growth to represent the out-of-Africa bottleneck and subsequent 
rapid expansion. YAML files representing the models and parameters in this 
section can be found in the linked GitHub repository.

We next stitched these one-population models together and refitted the 
resulting demographic model to get a two-population model 
that included continuous symmetric 
migration between MSL and GBR and an ancestral size expansion. Following this,
we added a Neandertal branch and fitted the 
Neandertal-modern human divergence time, Neandertal effective size, and the 
Neandertal-to-GBR admixture proportion. We also fitted the divergence time 
between the sampled Vindija Neandertal lineage and the Neandertal lineage which 
admixed with modern non-African humans. Finding that this parameter tended to 
converge to its lower bound (imposed by the estimated time of death of the 
Vindija Neandertal, ~55 kya), we fixed it to a value supported by literature 
\citep[90 kya,][]{prufer2017high} in a second round of optimization.
Rather than fitting it, we fixed the time of the Neandertal-GBR pulse to 50 kya.
Fixing parameters for which good estimates are available in the literature 
\citep[see e.g.,][]{sumer2025earliest} may be reasonable in the first stages of 
model optimization, and we expected this value to be especially difficult to fit 
due to the small pulse proportion and Neandertal sample size. In the second 
round of optimization on the three-population model, we refit all parameters 
aside from the two event times mentioned above to refine the estimates.

After this refitting step, we were left with parameters in good agreement with 
prior authors, with a Neandertal-modern human divergence time of 547 kya, 
MSL-GBR split time of 77 kya, and Neandertal-GBR pulse proportion of 2.7 
percent. We observe some large, systematic residuals in our best-fit model 
(Figure~\ref{fig:humans}G-I), suggesting model misspecification that could be
improved through further exploration of model space.

@@

The site frequency spectrum is just one of many options for inferring
evolutionary parameters from genetic data. It is straightforward to compute
from data, and there are many methods for predicting the SFS, making it an
appealing framework for demographic inference. As shown here in two examples,
we have incorporated the \demes format for specifying demographic models in
\moments to easily compute the SFS and perform demographic inference, including
computing uncertainties.

\moments is limited in the size of the dataset that it can consider. Sample
sizes can be large (in the hundreds), especially when considering models with
one or few populations, but will encounter memory issues when sample sizes are
extreme or as the number of populations grows. We are limited to a maximum of
five populations existing at any one time, including ancient samples. Other
SFS-based methods, such as \momi \citep{kamm2020efficiently, dilber2024faster},
can include additional populations, typically with smaller sample sizes.

When inferring models with multiple populations, especially four or more, both
the numbers of parameters and the space of plausible model topologies can grow
rapidly. It can therefore become challenging for the optimization routines to
find optima in parameter space, and overparameterization can become a problem.
One hint that models are overparameterized (or models misspecified) is runaway
behavior of parameter values, in which they converge to unreasonably small or
large values (or approach the imposed bounds). Starting with simple
parameterizations and sequentially adding complexity helps to alleviate this
issue. The challenge of large model space is more difficult to address, and
must be addressed through building and testing a wide range of plausible model
topologies. While time-consuming, a broad exploration of model space,
especially when considering many populations, is crucial for evaluating
robustness of particular features of inferred history.

Finally, natural selection distorts allele frequencies not only at sites
directly under selection, but also in linked regions of the genome through
background selection. This can have a sizeable impact on the SFS and lead to
biases in demographic inference \citep{ewing2016consequences, johri2021impact}.
In gene-sparse species, such as primates, considering data from regions away
from functionally constrained elements is feasible, as we did in example 2
here.  Alternatively, the genome may be partitioned by $B$-value \citep[a
measure of the local strength of background
selectionk,][]{mcvicker2009widespread} and regions experiencing low levels of
background selection may be used \citep[for an example,
see][]{medina2023demographic}. In some species with gene-dense genomes, such as
\emph{Drosophila} species, this may be infeasible, and interspersed putatively
neutral variation must be used, such synonymous variation or short introns,
which will still experience background selection. The appropriate strategy will
depend on the genome architecture of the species and genomic resources
available.

\section*{Online resources} \moments is available through \texttt{pypi} as
\texttt{moments-popgen}, and at \url{https://github.com/MomentsLD/moments} with
documentation at \url{https://momentsld.github.io/moments/}. The detailed
online examples accompanying this chapter can be found at
\url{https://github.com/StatisticalPopulationGenomics-2ndEd/moments/}.
